Mergeable Types

### 2017-02-25 paper/model.tex
edits

In this section, we describe the \name programming model through the
example of a collaborative drawing application we call \drawsome.

Fig.~\ref{fig:canvas-sig} shows the signature of the \drawsome
application. \drawsome represents a free-hand drawing canvas in terms
of a tree of quadrants.  Each quadrant is simply a leaf node
containing a single pixel (\C{r},\C{g}, or \C{b}) or a tree, if the
quadrant contains multiple pixels of different colors. Quadrants are
expanded into a tree structures as and when pixels are colored.  The
representation is thus optimized for sparse canvases, such as
whiteboards. The application supports three simple operations:
creating a new canvas, setting the pixel at a specified coordinate,
and returning the pixel at a given coordinate. The \C{merge} function
and \C{deriving versioned} directive are explained below.

\begin{figure}
\centering
\includegraphics[scale=0.5]{Figures/canvas-merging}

\caption{Merging concurrent versions (\C{v1} and \C{v2} ) of a drawing
canvas. \C{lca} is their common ancestor.}
\label{fig:canvas-merging}
\end{figure}

\drawsome lets multiple users collaborate on a canvas that is
conceptually shared among them. Under a shared-memory abstraction,
there would be a single copy of the canvas that is updated
concurrently by multiple clients; from the perspective of any single
client, the canvas could change without any explicit
intervention. \name ascribes functional semantics to sharing by
letting each client work on its own version of the state (the tree
data structure), later merging concurrent versions on-demand.  Thus,
the primary artifact of the \name programming model is a versioned
data structure in which different versions are managed by different
clients.

@@

\name requires a three-way \C{merge} function to merge the concurrent
versions of a drawing canvas (see Fig.~\ref{fig:merge-canvas}). The
three arguments include two concurrent versions (\C{v1} and \C{v2}),
and their least common ancestor (\C{lca}) - the version from which the
two concurrent versions evolved independently. The merge function can
make use of the pixel values of the common ancestor to merge the pixel
values on both the canvases. For instance, if the color of a pixel in
\C{v1} is white, and in \C{v2} it is green, and its color in \C{lca}
is white, then it means that only \C{v2} modified the color. Hence the
pixel is colored green in the merged canvas. On the other hand, if the
pixel is red in \C{v1}, then it means that both \C{v1} and \C{v2} have
modified the color. In such case, an appopriate color-mixing algorithm
can be used to determine the color of pixel.  For instance, the pixel
can be colored yellow - an additive combination of red and green. The
logic is illustrated in Fig.~\ref{fig:canvas-merging}.
\begin{figure}
\begin{center}
  \begin{ocaml}
let color_mix px1 px2 : pixel = 
let f = Char.code in
let h x y = Char.chr @@ (x + y)/ 2 in
let (r1,g1,b1) = (f px1.r, f px1.g, f px1.b) in
let (r2,g2,b2) = (f px2.r, f px2.g, f px2.b) in
let (r,g,b) = (h r1 r2, h g1 g2, h b1 b2) in {r=r; g=g; b=b}

@@

let rec merge lca v1 v2 = 
  if v1=v2 then v1
  else if v1=lca then v2
  else if v2=lca then v1
  else match (lca,v1,v2) with
    (*
     * The first three rules isomorphize lca, v1 and v2.
     *)
    | (_, B _, N px2) -> merge lca v1 @@ b_of_n px2
    | (_, N px1, B _) -> merge lca (b_of_n px1) v2
    | (N px, B _, B _) -> merge (b_of_n px) v1 v2
    | (B x, B x1, B x2) ->
        let tl_t' = merge x.tl_t x1.tl_t x2.tl_t in
        let tr_t' = merge x.tr_t x1.tr_t x2.tr_t in
        let bl_t' = merge x.bl_t x1.bl_t x2.bl_t in
        let br_t' = merge x.br_t x1.br_t x2.br_t in
          B {tl_t=tl_t'; tr_t=tr_t'; bl_t=bl_t'; br_t=br_t'}
    | (_, N px1, N px2) -> 
        (* pixels are merged by mixing colors *)
        let px' = color_mix px1 px2 in N px'
 \end{ocaml}
\caption{Merging different versions of a canvas.}
\label{fig:merge-canvas}
\end{center}
\end{figure}

The \name programming model lets programmers define and compose
concurrent computations around versioned data structures.
Fig.~\ref{fig:dali-monad} shows the signature of the \name module that
implements the programming model along the lines of the well-known
\C{State} monad~\cite{wadler-monad}. The monad encapsulates a
versioned functional state (\C{'a}) and the type \C{('a, 'b) t}
represents a monadic computation that returns a \C{'b} result.
Functions \C{return} and \C{bind} have their usual monadic
\begin{figure}
\begin{center}
  \begin{ocaml}
  module type DALI = sig
    type ('a, 'b) t
    val return : 'b -> ('a, 'b) t
    val bind : ('a, 'b) t -> ('b -> ('a, 'c) t) -> ('a, 'c) t
    val get_current_version: unit -> ('a, 'a) t
    val with_init_version_do: 'a -> ('a, 'b) t -> 'b
    val fork_version : ('a, 'b) t -> 'a unit t
    val sync_next_version: unit -> ?v:'a -> ('a, 'a) t
  end
  \end{ocaml}
\label{fig:dali-monad}
\caption{Signature of the \name monad}
\end{center}
\end{figure}
interpretation. \C{get\_current\_version} is like the \C{State}
monad's \C{get}; it returns the versioned state encapsulated by the
monad. \C{with\_init\_version\_do} runs a monadic computation against
a given initial version and returns the result. \C{fork\_version}
returns a computation that forks a new concurrent version from the
current version, and runs the given monadic computation asynchronously
against the forked version.  \C{sync\_next\_version} (simply called
\C{sync}) accepts a \emph{proposal} for the next version of the state;
this proposal is the current local version of the state that reflects
local modifications not yet witnessed by any other concurrently
executing computation.  The operation returns (via a monad) the actual
next version, which becomes the current version for the rest of the
computation.  This version is created by merging the proposal with a
subset of concurrent versions that have become available since the
last merge or fork. Thus, \C{sync} effectively lets a computation sync
with a subset of concurrent computations to obtain their latest
updates.

\begin{figure}
\centering
\begin{tabular}{l||l||l}
\begin{ocaml}
let alice_f : C.t unit t = 
  get () >>= fun c0 -> 
  fork bob_f >>= fun () ->
  let c0' = C.draw_line c0 
    {x=0;y=0}
    {x=4;y=0} in
  sync () ~v:c0' >>= 
  fun c1 ->
  let c1' = C.draw_line c1 
    {x=0;y=4} 
    {x=4;y=4} in
  sync () ~v:c1' >>= 
  fun c2 -> return ()
\end{ocaml}
&
\begin{ocaml}
let bob_f : C.t unit t = 
  get () >>= fun c0 -> 
  fork cheryl_f >>= 
  fun () ->
  let c0' = C.draw_line c0 
    {x=0;y=0} 
    {x=0;y=4} in
  sync () ~v:c0' >>= 
  fun c1 -> sync () >>= 
  fun c2 -> return ()
\end{ocaml}
&
\begin{ocaml}
let cheryl_f : C.t unit t = 
  get () >>= fun c0 -> 
  let c0' = C.draw_line c0 
    {x=4;y=0} 
    {x=4;y=0} in
  sync () ~v:c0' >>= 
  fun c1 -> sync () >>= 
  fun c2 -> return ()
\end{ocaml}
\\
\end{tabular}
\caption{\drawsome: A collaborative drawing session between Alice,
Bob, and Cheryl}
\label{fig:canvas-sessions-code}
\end{figure}

Fig.~\ref{fig:canvas-sessions-code} demonstrates how a collaborative
drawing session between Alice, Bob and Cheryl can be composed using
\name. A possible execution of the session is visualized in
Fig.~\ref{fig:canvas-sessions}. Assume that the session starts with
Alice on a $5\times 5$ blank canvas, as shown below:
\begin{ocaml}
  module C = Canvas;; 
  with_init_version_do (C.new_blank 5 5) alice_f 
\end{ocaml}
Alice starts by reading the current version of the canvas, which is
blank. She then invites Bob for collaboration by forking a new
concurrent version for Bob. Bob, in turn, invites Cheryl for
\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/canvas-sessions}
\caption{\drawsome: Collaborative drawing session visualized}
\label{fig:canvas-sessions}
\end{figure}
collaboration. All three of them start working on the blank canvas.
Alice draws a red horizontal line from $(0,0)$ (top-left) to $(4,0)$
(top-right) using \C{C.draw\_line}.\footnote{Its definitionis not
  shown, but can be constructed using \C{set\_px}} Meanwhile, Bob
draws a green vertical line from $(0,0)$ to $(0,4)$, and Cheryl draws
a similar line from $(4,0)$ to $(4,4)$. All three of them call
\C{sync} with their respective proposals ($C_0'$). While any partial
ordering of concurrent \C{syncs} is valid, we consider a linear order
where Cheryl's \C{sync} happens first, followed by Bob's and then
Alice's.  Cheryl's \C{sync} does not find any concurrent versions,
hence installs the proposed version ($C_0'$) as the next version on
Cheryl's branch. Bob's \C{sync} finds Cheryl's $C_0'$ as a concurrent
version, and merges it with its proposal to produce the next version
$C_1$, which is then installed on Bob's branch.  The least common
ancestor (LCA) for this merge is the initial version ($C_0$), and the
two concurrent versions are Bob's $C_0'$ and Cheryl's $C_0'$. Next,
Alice's \C{sync} finds Cheryl's $C_0'$ and Bob's $C_1$ as concurrent
versions, and merges them successively with Alice's proposal. For the
first merge, the two concurrent versions are Alice's $C_0'$ and
Cheryl's $C_0'$, and the LCA is the initial version ($C_0$). The
result of this merge is installed as the next version ($\C{V_1}$) on
Alice's branch. For the next merge, the two concurrent versions are
Alice's $V_1$ and Bob's $C_1$ and the LCA is Cheryl's
$C_0'$\footnote{Thus, the LCA of versions on two branches can lie
  outside both the branches.}. The result ($C_1$) becomes the next
version on Alice's branch, and the return value of Alice's \C{sync}.
Next, Alice draws a red horizontal line from $(0,4)$ to $(4,4)$,
proposes this canvas (\C{c1'} in the first column of
Fig.~\ref{fig:canvas-sessions-code}) as the next version to \C{sync}.
Since there are no concurrent versions, $C_1'$ becomes her next
version. The subsequent \C{sync} operations from Bob and Cheryl
propose no new versions, hence simply obtain Alice's $C_1'$ as next
versions.

The \drawsome example demonstrates the utility of mergeable data types
in building concurrent applications with conceptual sharing of
state. This example, however, implicitly makes certain assumptions
about the model and the underlying system, such as the existence of a
single least common ancestor (LCA) for any pair of versions, the
ability to access a previous version on any branch, and mergeability
of any two concurrent versions. Enforcing these guarantees in a fully
decentralized distributed setting requires addressing non-trivial
theoretical challenges described in the next section. Unlike the
canvas, the process of writing merge functions for data types with
non-trivial invariants could be quite involved. The following sections
describe a principled approach to deriving merge functions for any
data type given its invariants. Like \C{Canvas.merge}, merge functions
often need to compare multiple versions of a data type for structural
equality, which, when performed naively, incurs cost linear in the
size of versions. The representation of a versioned data type in \name
is optimized for performing comparisons that allows structural
equality to be determined in constant time. Such engineering details,
which contribute to the practicality of our programming model are also
discussed in subsequent sections.

>>>>

In this section, we describe the \name programming model using a
collaborative drawing application we call \drawsome as a motivating
example.

Fig.~\ref{fig:canvas-sig} shows the signature of the \drawsome
application. \drawsome represents a free-hand drawing canvas in terms
of a tree of quadrants.  Each quadrant is simply a leaf node
containing a single pixel (\C{r},\C{g}, or \C{b}) or a tree, if the
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.4]{Figures/canvas-merging}
\caption{Merging concurrent versions (\C{v1} and \C{v2}) of a drawing
canvas. \C{lca} is their common ancestor.}
\label{fig:canvas-merging}
\end{wrapfigure}
quadrant contains multiple pixels of different colors. Quadrants are
expanded into tree structures lazily as and when pixels are colored.
The representation is thus optimized for sparse canvases, such as
whiteboards. The application supports three simple operations:
creating a new canvas, setting the pixel at a specified coordinate,
and returning the pixel at a given coordinate. 
\drawsome lets multiple users collaborate on a canvas that is
conceptually shared among them. Under a shared-memory abstraction,
there would be a single copy of the canvas that is updated
concurrently by multiple clients; from the perspective of any single
client, the canvas could change without any explicit
intervention. \name ascribes functional semantics to sharing by
letting each client work on its own version of the state (the tree
data structure), later merging concurrent versions on-demand.  Thus,
the primary artifact of the \name programming model is a versioned
data structure in which different versions are managed by different
clients.

@@

\name requires a three-way \C{merge} function to merge the concurrent
versions of a drawing canvas (see Fig.~\ref{fig:merge-canvas}). The
three arguments include two concurrent versions (\C{v1} and \C{v2}),
and their least common ancestor (\C{lca}) - the version from which the
two concurrent versions evolved independently. The merge function can
make use of the pixel values of the common ancestor to merge the pixel
values on both the canvases. For instance, if the color of a pixel in
\C{v1} is white, and the same pixel in \C{v2} is green, and its color
in \C{lca} is white, then it means that only \C{v2} modified the
color. Hence the pixel is colored green in the merged canvas. On the
other hand, if the pixel is red in \C{v1}, then it means that both
\C{v1} and \C{v2} have modified the color. In such case, an
appropriate color-mixing algorithm can be used to determine the color
of pixel.  For instance, the pixel can be colored yellow - an additive
combination of red and green. The logic is illustrated in
Fig.~\ref{fig:canvas-merging}.  While the \drawsome\ merge operation
necessarily reflects domain-specific knowledge about the application,
it makes no reference to distribution or replication, and does not
impact the definition of any other operation, whose algorithms can be
developed for a sequential execution environment and seamlessly
transplanted to a distributed one.
\begin{figure}
\begin{center}
  \begin{ocaml}
let color_mix px1 px2 : pixel = 
let f = Char.code in
let h x y = Char.chr @@ (x + y)/ 2 in
let (r1,g1,b1) = (f px1.r, f px1.g, f px1.b) in
let (r2,g2,b2) = (f px2.r, f px2.g, f px2.b) in
let (r,g,b) = (h r1 r2, h g1 g2, h b1 b2) in {r=r; g=g; b=b}

@@

let rec merge lca v1 v2 = 
  if v1=v2 then v1
  else if v1=lca then v2
  else if v2=lca then v1
  else match (lca,v1,v2) with
    | (_, B _, N px2) -> merge lca v1 @@ b_of_n px2
    | (_, N px1, B _) -> merge lca (b_of_n px1) v2
    | (N px, B _, B _) -> merge (b_of_n px) v1 v2
    | (B x, B x1, B x2) ->
        let tl_t' = merge x.tl_t x1.tl_t x2.tl_t in
        let tr_t' = merge x.tr_t x1.tr_t x2.tr_t in
        let bl_t' = merge x.bl_t x1.bl_t x2.bl_t in
        let br_t' = merge x.br_t x1.br_t x2.br_t in
          B {tl_t=tl_t'; tr_t=tr_t'; bl_t=bl_t'; br_t=br_t'}
    | (_, N px1, N px2) -> 
        (* pixels are merged by mixing colors *)
        let px' = color_mix px1 px2 in N px'
 \end{ocaml}
\caption{Merging different versions of a canvas.}
\label{fig:merge-canvas}
\end{center}
\end{figure}
Fig.~\ref{fig:dali-monad} shows the signature of the \name module that
implements the programming model along the lines of the well-known
\C{State} monad~\cite{wadler-monad}. The monad encapsulates a
versioned functional state (\C{'a}) and the type \C{('a, 'b) t}
represents a monadic computation that returns a \C{'b} result.
Functions \C{return} and \C{bind} have their usual monadic
\begin{figure}
\begin{center}
  \begin{ocaml}
  module type DALI = sig
    type ('a, 'b) t
    val return : 'b -> ('a, 'b) t
    val bind : ('a, 'b) t -> ('b -> ('a, 'c) t) -> ('a, 'c) t
    val get_current_version: unit -> ('a, 'a) t
    val with_init_version_do: 'a -> ('a, 'b) t -> 'b
    val fork_version : ('a, 'b) t -> 'a unit t
    val sync_next_version: unit -> ?v:'a -> ('a, 'a) t
  end
  \end{ocaml}
  \caption{Signature of the \name monad}
  \label{fig:dali-monad}
\end{center}
\end{figure}
interpretation. \C{get\_current\_version} is like the \C{State}
monad's \C{get}; it returns the versioned state encapsulated by the
monad. \C{with\_init\_version\_do} runs a monadic computation against
a given initial version and returns the result. \C{fork\_version}
returns a computation that forks a new concurrent version from the
current version, and runs the given monadic computation asynchronously
against the forked version.  \C{sync\_next\_version} (simply called
\C{sync}) accepts a \emph{proposal} for the next version of the state;
this proposal is the current local version of the state that reflects
local modifications not yet witnessed by any other concurrently
executing computation.  The operation returns (via a monad) the actual
next version, which becomes the current version for the rest of the
computation.  This version is created by merging the proposal (the
local state) with a subset of concurrent versions that have become
available since the last merge or fork.  Thus, \C{sync} effectively
lets a computation sync with all computations operating over replicas
of the same state that are available (i.e., reachable in the
network)\footnote{The implementation implicitly triggers a sync
  whenever a detected network partition is repaired.}  to obtain their
latest updates.
\begin{figure}
\centering
\begin{tabular}{l||l||l}
\begin{ocaml}
let alice_f : C.t unit t = 
  get () >>= fun c0 -> 
  fork bob_f >>= fun () ->
  let c0' = C.draw_line c0 
    {x=0;y=0}
    {x=4;y=0} in
  sync () ~v:c0' >>= 
  fun c1 ->
  let c1' = C.draw_line c1 
    {x=0;y=4} 
    {x=4;y=4} in
  sync () ~v:c1' >>= 
  fun c2 -> return ()
\end{ocaml}
&
\begin{ocaml}
let bob_f : C.t unit t = 
  get () >>= fun c0 -> 
  fork cheryl_f >>= 
  fun () ->
  let c0' = C.draw_line c0 
    {x=0;y=0} 
    {x=0;y=4} in
  sync () ~v:c0' >>= 
  fun c1 -> sync () >>= 
  fun c2 -> return ()
\end{ocaml}
&
\begin{ocaml}
let cheryl_f : C.t unit t = 
  get () >>= fun c0 -> 
  let c0' = C.draw_line c0 
    {x=4;y=0} 
    {x=4;y=0} in
  sync () ~v:c0' >>= 
  fun c1 -> sync () >>= 
  fun c2 -> return ()
\end{ocaml}
\\
\end{tabular}
\caption{\drawsome: A collaborative drawing session between Alice,
Bob, and Cheryl}
\label{fig:canvas-sessions-code}
\end{figure}

Fig.~\ref{fig:canvas-sessions-code} demonstrates how a collaborative
drawing session between Alice, Bob and Cheryl can be composed using
\name. A possible execution of the session is visualized in
Fig.~\ref{fig:canvas-sessions}. Assume that the session starts with
Alice on a $5\times 5$ blank canvas, as shown below:
\begin{ocaml}
  module C = Canvas;; 
  with_init_version_do (C.new_blank 5 5) alice_f 
\end{ocaml}
Alice starts by reading the current version of the canvas, which is
blank. She then invites Bob for collaboration by forking a new
concurrent version for Bob. Bob, in turn, invites Cheryl.
\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/canvas-sessions}
\caption{\drawsome: Collaborative drawing session visualized}
\label{fig:canvas-sessions}
\end{figure}
All three start working on the blank canvas on different replicas.
Alice draws a red horizontal line from $(0,0)$ (top-left) to $(4,0)$
(top-right) using \C{C.draw\_line}.\footnote{Its definition is not
  shown, but can be constructed using \C{set\_px}} Meanwhile, Bob
draws a green vertical line from $(0,0)$ to $(0,4)$, and Cheryl draws
a similar line from $(4,0)$ to $(4,4)$. All three of them call
\C{sync} with their respective proposals ($C_0'$). While any partial
ordering of concurrent \C{syncs} is valid, we consider a linear order
where Cheryl's \C{sync} happens first, followed by Bob's and then
Alice's.  Cheryl's \C{sync} does not find any concurrent versions,
hence installs the proposed version ($C_0'$) as the next version on
Cheryl's branch. Bob's \C{sync} finds Cheryl's $C_0'$ as a concurrent
version, and merges it with its proposal to produce the next version
$C_1$, which is then installed on Bob's branch.  The least common
ancestor (LCA) for this merge is the initial version ($C_0$), and the
two concurrent versions are Bob's $C_0'$ and Cheryl's $C_0'$. Next,
Alice's \C{sync} finds Cheryl's $C_0'$ and Bob's $C_1$ as concurrent
versions, and merges them successively with Alice's proposal. For the
first merge, the two concurrent versions are Alice's $C_0'$ and
Cheryl's $C_0'$, and the LCA is the initial version ($C_0$). The
result of this merge is installed as the next version ($\C{V_1}$) on
Alice's branch. For the next merge, the two concurrent versions are
Alice's $V_1$ and Bob's $C_1$ and the LCA is Cheryl's
$C_0'$\footnote{Thus, the LCA of versions on two branches can lie
  outside both the branches.}. The result ($C_1$) becomes the next
version on Alice's branch, and the return value of Alice's \C{sync}.
Next, Alice draws a red horizontal line from $(0,4)$ to $(4,4)$,
proposes this canvas (\C{c1'} in the first column of
Fig.~\ref{fig:canvas-sessions-code}) as the next version to \C{sync}.
Since there are no concurrent versions, $C_1'$ becomes her next
version. The subsequent \C{sync} operations from Bob and Cheryl
propose no new versions, hence simply obtain Alice's $C_1'$ as next
versions.

The \drawsome example demonstrates the utility of mergeable data types
in building concurrent applications with conceptual sharing of
state. This example, however, implicitly makes certain assumptions
about the model and the underlying system, such as the existence of a
single least common ancestor (LCA) for any pair of versions, the
ability to access a previous version on any branch, and mergeability
of any two concurrent versions. Enforcing these guarantees in a fully
decentralized distributed setting requires addressing non-trivial
theoretical challenges described in
Section~\ref{sec:semantics}. Unlike the canvas, the process of writing
merge functions for data types with non-trivial invariants could be
quite involved. The following section describes a principled approach
to deriving merge functions for any data type given its
invariants. Like \C{Canvas.merge}, merge functions often need to
compare multiple versions of a data type for structural equality,
which, when performed naively, incurs cost linear in the size of
versions. The representation of a versioned data type in \name is
optimized for performing comparisons that allows structural equality
to be determined in constant time. Such engineering details, which
contribute to the practicality of our programming model, are discussed
in Sec.~\ref{sec:impl}.

### 2017-02-25 paper/semantics.tex
System Model
edits

We formalize our ideas in the context of a lambda calculus ($\lang$)
shown in Fig.~\ref{fig:opsem}. Expressions of $\lang$ are variables,
constants, and \name primitives composed using the lambda combinator.
For brevity, we use short names for \name primitives: \C{run} for
\C{with\_init\_version\_do}, and \C{fork} for \C{fork\_version}. To
simplify the technical development, \name's \C{sync\_next\_version}
operation is broken down into two primitives - \C{push} and \C{pull},
which can be lambda-composed get the desired effect:
\begin{smathpar}
  \C{sync}\;x \;=\; (\lambda y.\pull)\; (\push\,x)
\end{smathpar}
The semantics of\C{get\_current\_version} is subsumed by \C{pull},
hence elided.  Values ($v$) are constants and lambda abstractions.  A
program ($p$) is a parallel composition of threads, where each thread
is an expression ($s$) indexed by the corresponding thread identifier
($t$).

Fig.~\ref{fig:opsem} also shows the syntax of \emph{branches}, which
are the artifacts of evaluation and only appear during the run-time. A
branch is a non-empty sequence of tagged values, where the tag
captures the abstract run-time operation that led to the creation of
the value. It is implicitly assumed that each value added to a branch
is uniquely identifiably, hence no two values on a branch are equal.
The uniqueness assumption is later extended to a collection of
branches that constitute a branching structure. A real implementation
meets this assumption by versioning values across the branches. Thus,
in reality, branches contain \emph{versions} which denote values. The
semantics, however, doesn't make this distinction, and uses values and
versions interchangeably.

Small-step operational semantics of $\lang$ is defined via reduction
relation ($\stepsto$) that relates \emph{program states}. A program
state ($p;\,H$) consists of a program $p$ and a \emph{branch history}
$H$ that maps thread Ids to corresponding branches; each thread is
associated with a branch during the evaluation. Evaluation contexts
have been defined separately for expressions ($E$) and programs ($P$),
with the latter subsuming the former. $E$ is defined to evaluate the
first argument of a \C{run} expression to a value that constitutes the
initial version (recall that \C{run} models \name's
\C{with\_init\_version\_do}). Program evaluation context
non-deterministically picks one of the threads to evaluate. The admin
rule that relates transitions of holes to transitions of expressions
and programs is straightforward, hence elided. Rest of the reduction
rules are presented in Fig.~\ref{fig:opsem}. For brevity, we write
$H(t\mapsto (v,f))$ to denote the proposition that $H$ maps $t$ to
$(v,f)$. The notation $H[t \mapsto (v,f)]$ as usual denotes
the extension of $H$ with the binding $t \mapsto (v,f)$.

Reduction rules let expression evaluation take a step by rewriting the
expression and suitably updating the branch history ($H$). The
\rulelabel{E-Run} rule is applicable only when $H$ is empty, i.e.,
when no prior branching structure exists. The rule rewrites the
$\C{run}\;v\;s$ expression to $s$, while creating a new branching
structure with two branches: a \emph{top} branch that has just the
initial version (tagged with \C{INIT}), and a branch for the current
thread ($t$) forked-off from the top branch.  The first version on the
current branch ($H(t)$) denotes the same value ($v$) as the initial
version on the top branch, although versions themselves are deemed
distinct. The new version is tagged with a \C{FORK} tag that keeps the
record of its orgin, namely the \C{fork} operation and the branch from
which the current branch is forked. The \rulelabel{E-Fork} rule forks
a new thread with a fresh id ($t'$) and adds it to the thread pool.
The corresponding branch ($H(t')$) is forked from the parent thread's
branch ($H(t)$). The semantics of branch forking is same as described
above. The \C{fork} expression in the parent thread evaluates to
\C{()}. The \rulelabel{E-Push} rule creates a new version on the
current branch ($H(t)$) using the pushed value ($v$).
\rulelabel{E-App} is the standard beta reduction rule.

The semantics non-deterministically chooses \rulelabel{E-Pull} or
\rulelabel{E-Pull-Wait} rules to reduce a \C{pull} expression. The
\rulelabel{E-Pull} rule reduces \C{pull} to \C{()}, and returns the
latest version on the current branch. The \rulelabel{E-Pull-Wait} rule
can be thought of as a stutter step; it doesn't reduce \C{pull}, but
updates the branching structure by merging (the latest version of) a
concurrent branch ($H(t')$) into (the latest version of) the current
branch ($H(t)$), and extending the current branch with the merged
version ($v_m$). The new version is tagged with a \C{MERGE} tag that,
like a \C{FORK} tag, records its origin. The rule assumes a function
\C{lca} that computes the \emph{least common ancestor} (LCA) for the
latest versions on the given pair of branches. The semantics of LCA is
discussed below. The \rulelabel{E-Pull-Wait} and \rulelabel{E-Pull}
rules thus let a thread sync with a subset of concurrent threads in
multiple steps before returning the result of the \C{pull}. Since
\C{sync} is a composition of \C{push} and \C{pull}, its behavior can
be explained thus: \C{sync} pushes the given value onto the current
(local) branch, merges a (possibly empty) subset of concurrent
branches into the local branch, and returns the result.

@@

>>>>

We formalize our ideas in the context of a lambda calculus ($\lang$)
shown in Fig.~\ref{fig:opsem}. Expressions of $\lang$ are variables,
constants, and \name primitives composed using the lambda combinator.
For brevity, we use short names for \name primitives: \C{run} for
\C{with\_init\_version\_do}, and \C{fork} for \C{fork\_version}. To
simplify the technical development, \name's \C{sync\_next\_version}
operation is broken down into two primitives - \C{push} and \C{pull},
which can be composed to get the desired effect of the original:
\begin{smathpar}
  \C{sync}\;x \;=\; (\lambda y.\pull)\; (\push\,x)
\end{smathpar}
The semantics of\C{get\_current\_version} is subsumed by \C{pull},
hence elided.  Values ($v$) are constants and lambda abstractions.  A
program ($p$) is a parallel composition of threads, where each thread
is an expression ($s$) indexed by the corresponding thread identifier
($t$).

Fig.~\ref{fig:opsem} also shows the syntax of \emph{branches}, which
are artifacts of evaluation and only appear during run-time. A branch
is a non-empty sequence of tagged values, where the tag captures the
abstract run-time operation that led to the creation of the value. It
is implicitly assumed that each value added to a branch is uniquely
identifiable, and hence no two values on a branch are equal.  The
uniqueness assumption is later extended to a collection of branches
that constitute a branching structure. A real implementation meets
this assumption by versioning values across the branches. Thus, in
reality, branches contain \emph{versions} which denote values.  To
simplify the presentation, the semantics, does not make this
distinction, and uses values and versions interchangeably.

The small-step operational semantics of $\lang$ is defined via a
reduction relation ($\stepsto$) that relates \emph{program states}. A
program state ($p;\,H$) consists of a program $p$ and a \emph{branch
  history} $H$ that maps thread identifiers to corresponding branches;
each thread is associated with a branch during evaluation. Evaluation
contexts have been defined separately for expressions ($E$) and
programs ($P$), with the latter subsuming the former.

$E$ is defined to evaluate the first argument of a \C{run} expression
to a value that constitutes an initial version (recall that \C{run}
models \name's \C{with\_init\_version\_do}). A program evaluation
context non-deterministically picks an extant thread to evaluate. The
administrative rule that relates transitions of holes to transitions
of expressions and programs is straightforward, and hence elided. The
remaining core reduction rules are presented in
Fig.~\ref{fig:opsem}. For brevity, we write $H(t\mapsto (v,f))$ to
denote the proposition that $H$ maps $t$ to $(v,f)$. The notation $H[t
  \mapsto (v,f)]$ denotes the extension of $H$ with the binding $t
\mapsto (v,f)$, as usual.

Reduction rules let expression evaluation take a step by rewriting the
expression and suitably updating the branch history ($H$).
\rulelabel{E-App} is the standard beta reduction rule.  The
\rulelabel{E-Run} rule is applicable only when $H$ is empty, i.e.,
when no prior branching structure exists, and commences a distributed
computation with a corresponding version tree. The rule rewrites the
$\C{run}\;v\;s$ expression to $s$, while creating a new branching
structure with two branches: a \emph{top} branch that has just the
initial version (tagged with \C{INIT}), and a branch for the current
thread ($t$) forked-off from the top branch.  The first version on the
current branch ($H(t)$) denotes the same value ($v$) as the initial
version on the top branch, although versions themselves are deemed
distinct. The new version is tagged with a \C{FORK} tag that keeps the
record of its orgin, namely the \C{fork} operation and the branch from
which the current branch is forked. The \rulelabel{E-Fork} rule forks
a new thread with a fresh id ($t'$) and adds it to the thread pool.
The corresponding branch ($H(t')$) is forked from the parent thread's
branch ($H(t)$). The semantics of branch forking is the same as
described above. The \C{fork} expression in the parent thread
evaluates to \C{()}. The \rulelabel{E-Push} rule creates a new version
on the current branch ($H(t)$) using the pushed value ($v$).  Although
our semantics does not directly expose heap-allocated values, the
intention is that $v$ is a replicated object, manipulated on the local
heap that, after push, now becomes subject to merging and coordination
with other replicas.

The semantics non-deterministically chooses \rulelabel{E-Pull} or
\rulelabel{E-Pull-Wait} rules to reduce a \C{pull} expression. The
\rulelabel{E-Pull} rule reduces \C{pull} to \C{()}, and returns the
latest version on the current branch. The \rulelabel{E-Pull-Wait} rule
can be thought of as a stutter step; it doesn't reduce \C{pull}, but
updates the branching structure by merging (the latest version of) a
concurrent branch ($H(t')$) into (the latest version of) the current
branch ($H(t)$), and extending the current branch with the merged
version ($v_m$). The new version is tagged with a \C{MERGE} tag that,
like a \C{FORK} tag, records its origin. The rule assumes a function
\C{lca} that computes the \emph{least common ancestor} (LCA) for the
latest versions on the given pair of branches. The semantics of LCA is
discussed below. The \rulelabel{E-Pull-Wait} and \rulelabel{E-Pull}
rules thus let a thread synchronize with other distributed threads
manipulating versions of replicas in multiple steps before returning
the result of the \C{pull}. Since \C{sync} is a composition of
\C{push} and \C{pull}, its behavior can be explained thus: \C{sync}
pushes the given value onto the current (local) branch, merges a
(possibly empty) subset of concurrent branches into the local branch,
and returns the result.  (The operation is not guaranteed to
synchronize with all concurrent branches because not all such branches
may be available.)

@@

\input{system-model}
